### 2020-12-10 doc/manuscript/Dow_manuscript_Nature.Rmd
small changes

Thus, while it is increasingly clear that warming spring temperatures lengthen the growing season and modestly increase annual carbon sequestration, indirect observations of these effects - like those involving leaf phenology - are not enough to draw firm conclusions on the future of forests. Instead, direct observations of intraannual stem-growth should be compared between years with a variety of spring climate conditions so a clearer relationship between climate and stem-growth can be deciphered. Looking directly at the stem-growth is advantageous as it allows examination of individual trees as opposed to larger landscape scale patterns obtained through eddy-covariance or remote sensing studies. A direct, fine scale, individual based study would allow a look into the effect of warming springtime on stem-growth phenology and the related shifts in growing season length and $NEE$.

@@

Tree-ring records give some hints that warmer springs may increase annual growth ([@Helcoski_growing_2019]; probably some more), but tree-rings tell us nothing about the timing of that growth. Similarly, large scale observations of leaf-phenolgy and eddy-covariance also hint that warmer springs increase annual growth but tell us little about the maximum C uptake rates, or interspecific differences in total carbon sequestered.

## (We also know little about differences across species or functional groups, which matters because it will affect how different species fare under climate change )
Leaf and wood phenology traits, along with their apparent relationships, are dissimilar among trees of different wood types (ring-porous and diffuse-porous; **D'Orangeville et al. **). Interspecific variation of this sort likely stems from evolutionary differences and adaptations to local climates where these trees can be found. Ring-porous trees have large xylem cells which are prone to embolism during the winter. These cells are repaired before leaf development in the Spring, corresponding with an earlier initiaiton of stem-growth. Diffuse-porous xylem cells are smaller, and less prone to embolism during the Winter, meaning they do not need to be repaired in order to restore hydraulic functionality before beginning the leaf development process. Results from the paper by D’Orangeville et al. (in review) have shown that ring-porous trees reach 25 and 50% of their total yearly growth during spring, while diffuse-porous trees reach these milestones in the summer, but both reached 75% total annual growth around the same time. If leaf development is sped up when spring temperatures are warmer, we should seek to understand how this shift influences stem-growth timing and other phenological variables.

@@

We found that warmer spring temperatures in temperate deciduous forests cause earlier tree growth but have little effect on annual woody productivity. Using dendroband data, we found that the growing period is generally shifting back in the season as the critical srping windows are warming, but having little, or negative effect on peak growing season duration. Maximum growth rates and total seasonal growth were virtually unchanged when this period warmed, but the Day of year where the maximum growth rate occured shitfed back almost in unison with the DOY_50. The lack of influence Spring TMAX exudes on total season growth found with the dendroband data agrees with patterns observed in tree cores. Annual tree growth depends more on peak growing season conditions than early season advancement.

@@

~~~~

## Cam's previous content

We have shown that intraannual growth patterns in North American temperate trees change in response to increasing spring temperatures. The magnitude of these changes depends largely on the evolutionary differences inherent in the two wood types analyzed here. Ring-porous trees must repair winter embolism of their large xylem vessels before leaf expansion, achieving roughly 25% total yearly growth simultaneously with bud burst, and 50% yearly growth around the time of 75% leaf emergence (D'Orangeville et al,). On the other hand, diffuse-porous trees are less prone to cavitation and so do not need to repair winter embolism before bud burst, allowing up to 75% leaf out before reaching their 25% yearly growth totals (D'Orangeville et al). These differences are likely evolutionary hydraulic adaptations to climate (Baas & Wheeler, 2011). The presence of wide-xylem cells in ring-porous trees allows for greater water conductivity in the spring at the cost of increased cavitation risk during spring drought, and high likelihood of damage during frost events (Cochard & Tyree, 1990; Sperry et al, 1994). The pre-leaf stem growth occurring in these species is fueled by non-structural carbon stores, often used for other cell maintenance activities within the plant (Sperry et al, 1994; Palacio et al, 2014; Sevanto et al, 2014). The dependence on these carbon stores for cambial activation could prove to be detrimental to long-term health of the tree, as exhaustion of reserves leads to carbon starvation (Sevanto et al, 2014). Diffuse-porous trees have narrower xylem cells lowering the risk of cavitation, but leading to a decreased water conductivity in the spring, revealing a growth strategy more focused on a relatively faster, shorter growing season in the Summer (Sperry et al, 1994). This allows the diffuse-porous trees to conserve their carbon stores, making them potentially more resistant to pest/pathogen invasions?, but exposing them to higher temperatures and drought risk during their Summer-centered growth period.

The mechanisms controlling activation of cambium reactivation is still debated, as some speculate Auxin, the chemical believed to kick-start the reactivation process, originates in the leaves (Frankenstein et al, 2005). However, the fact that ring-porous species complete 25% growth before the presence of leaf buds indicates there may be another source of Auxin production. *READ MORE ABOUT THIS!* The relationship between earlier cambial reactivation and early spring temperature signals that the mechanism is influenced by temperature?

Under the warming regime analyzed here, ring-porous trees reach 25% and 50% growth milestones earlier (insert actual # days?) in the year, while timing of 75% growth !!! Changed by constant amount !!! . This implies that the growing season of ring-porous trees expanded due to spring temperatures, which are projected to continue to increase as anthropogenic global warming intensifies. In fact, we found a positive correlation between ring-porous growing season length and spring temperature, while no such relationship was observed in diffuse-porous trees. The earlier reactivation of cambial cells is potentially exposing these trees to higher risk of cavitation or damage during frost events (Cochard & Tyree, 1990). Damage of this nature would then need to be repaired using carbon reserves, which may be in short demand as they are already being used for initial growth.This would negatively impact the plant's ability to maintain cell health, or reduce the ability to respond to invasion of pests or pathogens. Alternatively, the extension of growth season and higher yearly growth totals like we see here, could give ring-porous trees a competitive advantage over their non-responsive diffuse-porous counterparts.

Diffuse-porous growth timing seems relatively unaffected by early spring temperatures, instead showing a higher correlation with late spring or Summer temps. This disconnect means that diffuse-porous trees will continue to focus their growing season in the Summer, which is projected to become less conducive to tree growth as temperatures and drought intensity increase in some areas. ---- Studies have found that diffuse-porous trees in the Eastern USA show a larger reduction in ring-width increment during, and following Summer droughts when compared to species of other wood types (Orwig & Abrams, 1997; Kannenberg et al, 2018) as well as greater sensitivity to Summer precipitation patterns (Elliot et al, 2015). The potential for more frequent reduction in growth coupled with the increased growing season length of ring-porous species may lead to an alteration in forest species composition as diffuse-porous species are outcompeted.

--the range of some of these species may migrate North to cooler summer conditions becoming potentially extant in the Southern / mid Atlantic of the USA.--    (Is mortality of ring-porous species higher in recent decades? Hard to pinpoint, but would likely be caused by warmer early-spring followed by late frost events exhausting NSC reserves?)

This analysis is one of a first generation of intraannual dendroband tree growth studies, and as such, is only the tip of the ice berg. Here, we focused our analysis purely on early spring warming. This falls short of the reality of our changing climate, where each season is being affected in some way, leading to a combination of stressors on plant growth. For example, diffuse porous trees seemed to show little, if any, relationship with the climate window defined here, but if we shift that window forward 20 days, the effect increases dramatically, signalling a greater importance of May/June $T_{MAX}$ on trees of that wood type (figure or table?). This revelation agrees with the results from other climate sensitivity studies like Helcoski et al (2019), which found differences in interspecific growth responses to several climate variables throughout the year. Additionally, Other studies have found that Autumn leaf senescence is being delayed in response to warmer temperatures in the previous year's Winter or Autumn (), positive correlation between growth and higher winter temperatures (Heide, 1993; Delpierre et al, 2009), stunted total productivity induced by drought and high $T_{MAX}$ in the current year's Summer (), and an increase in diffuse-porous tree growth during the growing season following an abnormally cold winter (Benito & Pederson, 2015; Elliot et al, 2015). How these relationships may impact intraannual growth patterns is uncertain, but identifying such relationships will help us further understand how climate change is impacting forests.

We present a stepping stone for future studies to build upon.

>>>>

@@

Tree-ring records give some hints that warmer springs may increase annual growth ([@Helcoski_growing_2019]; probably some more), but tree-rings tell us nothing about the timing of that growth. On the other hand, earlier leaf development shows us how the growing season is advancing, but can't tell us about the total carbon sequestered, while neither can be used to map out intra-annual growth patterns.

Thus, while it is increasingly clear that warming spring temperatures lengthen the growing season and modestly increase annual carbon sequestration, current methods do not allow firm conclusions on the future of forests.A combination of existing methods - tree cores and remote sensing of leaf-phenology traits - should be supplemented by dendrometer band data which allows for the modeling of intra-annual growth patterns in individuals. A mixture of these three would produce the most complete picture of spring climate influence on forests.

## (We also know little about differences across species or functional groups, which matters because it will affect how different species fare under climate change )
*(Too much detail here? how about just keeping some variation of the first and last sentences)*
Leaf and wood phenology traits, along with their apparent relationships, are dissimilar among trees of different wood types (ring-porous and diffuse-porous; **D'Orangeville et al. **). Interspecific variation of this sort likely stems from evolutionary differences and adaptations to local climates where these trees can be found. Ring-porous trees have large xylem cells which are prone to embolism during the winter. These cells are repaired before leaf development in the Spring, corresponding with an earlier initiaiton of stem-growth. Diffuse-porous xylem cells are smaller, and less prone to embolism during the Winter, meaning they do not need to be repaired in order to restore hydraulic functionality before beginning the leaf development process. Results from the paper by D’Orangeville et al. (in review) have shown that ring-porous trees reach 25 and 50% of their total yearly growth during spring, while diffuse-porous trees reach these milestones in the summer, but both reached 75% total annual growth around the same time. If leaf development is sped up when spring temperatures are warmer, we should seek to understand how this shift influences stem-growth timing and other phenological variables.

@@

We found that warmer spring temperatures in temperate deciduous forests cause earlier tree growth but have little effect on annual woody productivity. Using dendroband data, we found that the peak growing season is generally advancing as the critical spring windows are warming but having little, or negative effect, on $L_{pgs}$. $g_{max}$ and $\Delta DBH$ were virtually unchanged when this period warmed, but $DOY_{g_{max}}$ shifted back almost in unison with $DOY_{50}$. The lack of influence Spring TMAX exudes on $\Delta DBH$ found with the dendroband data agrees with patterns observed in tree cores. Annual tree growth depends more on peak growing season conditions than early season advancement.

@@
